Cooperative Kernels: Safe Multitasking for Blocking Algorithms on GPUs

### 2016-11-13 write-ups/PLDI/paper.tex
Shortening intro.

\paragraph{The needs of irregular data-parallel algorithms}
Acceleration of general-purpose computations on graphics processing
units (GPUs) has tended to focus on \emph{regular} data-parallel
algorithms, for which work to be processed can be evenly split between
the workgroups that execute a GPU kernel ahead of time.  However, many
interesting data-parallel algorithms are \emph{irregular}: the amount
of work to be processed is unknown ahead of time and may be determined
by the computation itself.  The last decade has seen growing interest in accelerating irregular
data-parallel algorithms on GPUs, particularly algorithms that
traverse and manipulate linked data structures~\cite{owens-persistent,TPO10,DBLP:conf/ipps/KaleemVPHP16,DBLP:conf/ipps/DavidsonBGO14,DBLP:conf/hipc/HarishN07,DBLP:journals/topc/MerrillGG15,DBLP:conf/egh/VineetHPN09,DBLP:conf/ppopp/NobariCKB12,DBLP:conf/hpcc/SolomonTT10a,DBLP:conf/popl/PrabhuRMH11,DBLP:conf/ppopp/Mendez-LojoBP12,DBLP:conf/oopsla/PaiP16,DBLP:conf/ipps/KaleemVPHP16,DBLP:conf/oopsla/SorensenDBGR16}.  \ADComment{Tyler, can you add relevant workstealing papers if not here already?}

Irregular algorithms usually require \emph{blocking synchronization}
between workgroups, to balance load and communicate intermediate
results.  For example, many graph algorithms employ a level-by-level strategy, requiring a global barrier between each level.
As another example, work stealing algorithms require each workgroup
to maintain a queue, which is typically protected by a mutex to enable
stealing by other workgroups.

To operate correctly, a blocking concurrent algorithm requires
\emph{fair} scheduling of threads.  In the context of irregular
data-parallel algorithms accelerated on a GPU, this translates to
requiring fair scheduling of workgroups.  Without fairness, a
workgroup may fail to make progress, leading to starvation.  For
example, if workgroup $A$ holds a mutex, an unfair scheduler may cause
another workgroup $B$ to spin-wait forever for the mutex to be
released.  Similarly, an unfair scheduler can cause a workgroup to spin-wait
indefinitely at a global barrier so that other workgroups do not make progress
towards the barrier.

\paragraph{A degree of fairness: occupancy-bound execution}
GPU schedulers are \emph{not} fair. the current GPU
programming models---OpenCL~\cite{opencl2Spec}, \ADComment{Tyler, please add best CUDA since I've asked you to add it below in any case} CUDA~\cite{...}  and
HSA~\cite{HSAprogramming11}---specify almost no guarantees regarding scheduling of
workgroups, and implementations of these programming models do not
provide fair scheduling in practice.  Roughly speaking, each workgroup
executing a GPU kernel is mapped to a hardware \emph{compute
  unit}.\footnote{In practice the situation can be more complex: if a
  kernel is sufficiently simple and if workgroups are small enough,
  multiple workgroups may be mapped to a single compute unit.  We assume, for the purpose of the current discussion, that the kernel is sufficiently complex that this does not happen.}
%
If a kernel is executed with more workgroups than there are compute
units then at least two workgroups must share a single compute unit.
The simplest way for a GPU driver to handle this is via an \emph{occupancy-bound}
execution model~\cite{DBLP:conf/oopsla/SorensenDBGR16}. \ADComment{Tyler, please add best cite for OB model; I guess our OOPSLA paper introduced the term, but we should cite that + Owens?}  With this model, once a workgroup has commenced execution on a
compute unit (it has become \emph{occupant}), the workgroup has
exclusive access to the compute unit until it has finished execution,
after which a subsequent workgroup can be bound to the compute unit.
Experiments suggest that the occupancy-bound execution model is widely
implemented by today's GPU devices and drivers~\cite{DBLP:conf/oopsla/SorensenDBGR16}. \ADComment{For the previous cite, should we cite other papers as well as ours?}  It is
evidently a simple model for device vendors to implement efficiently.

>>>>

\paragraph{The needs of irregular data-parallel algorithms}
%Acceleration of general-purpose computations on graphics processing
%units (GPUs) has tended to focus on \emph{regular} data-parallel
%algorithms, for which work to be processed can be evenly split between
%the workgroups that execute a GPU kernel ahead of time.  However,
Many
interesting data-parallel algorithms are \emph{irregular}: the amount
of work to be processed is unknown ahead of time and may be determined
by the computation itself.  The last decade has seen growing interest in accelerating irregular
data-parallel algorithms on GPUs, particularly algorithms that
traverse and manipulate linked data structures~\cite{owens-persistent,TPO10,DBLP:conf/ipps/KaleemVPHP16,DBLP:conf/ipps/DavidsonBGO14,DBLP:conf/hipc/HarishN07,DBLP:journals/topc/MerrillGG15,DBLP:conf/egh/VineetHPN09,DBLP:conf/ppopp/NobariCKB12,DBLP:conf/hpcc/SolomonTT10a,DBLP:conf/popl/PrabhuRMH11,DBLP:conf/ppopp/Mendez-LojoBP12,DBLP:conf/oopsla/PaiP16,DBLP:conf/ipps/KaleemVPHP16,DBLP:conf/oopsla/SorensenDBGR16}.  \ADComment{Tyler, can you add relevant workstealing papers if not here already?}

Irregular algorithms usually require \emph{blocking synchronization}
between workgroups, to balance load and communicate intermediate
results.  Many graph algorithms employ a level-by-level strategy, requiring a global barrier between each level.
Work stealing algorithms require each workgroup
to maintain a queue, which is typically protected by a mutex to enable
stealing by other workgroups.

To operate correctly, a blocking concurrent algorithm requires
\emph{fair} scheduling, which in the context of GPUs means fair scheduling of workgroups.  Without fairness, a
workgroup may fail to make progress, leading to starvation.  For
example, if workgroup $A$ holds a mutex, an unfair scheduler may cause
another workgroup $B$ to spin-wait forever for the mutex to be
released.  Similarly, an unfair scheduler can cause a workgroup to spin-wait
indefinitely at a global barrier so that other workgroups do not make progress
towards the barrier.

\paragraph{A degree of fairness: occupancy-bound execution}
GPU schedulers are \emph{not} fair. the current GPU
programming models---OpenCL~\cite{opencl2Spec}, \ADComment{Tyler, please add best CUDA since I've asked you to add it below in any case} CUDA~\cite{...}  and
HSA~\cite{HSAprogramming11}---specify almost no guarantees regarding scheduling of
workgroups, and current GPUs do not
provide fair scheduling in practice.  Roughly speaking, each workgroup
executing a GPU kernel is mapped to a hardware \emph{compute
  unit}.\footnote{In practice the situation can be more complex: if a
  kernel is sufficiently simple and if workgroups are small enough,
  multiple workgroups may be mapped to a single compute unit.  We assume, for the purpose of the current discussion, that the kernel is sufficiently complex that this does not happen.}
%
If a kernel is executed with more workgroups than there are compute
units then at least two workgroups must share a single compute unit.
The simplest way for a GPU driver to handle this is via an \emph{occupancy-bound}
execution model~\cite{DBLP:conf/oopsla/SorensenDBGR16}. \ADComment{Tyler, please add best cite for OB model; I guess our OOPSLA paper introduced the term, but we should cite that + Owens?}  With this model, once a workgroup has commenced execution on a
compute unit (it has become \emph{occupant}), the workgroup has
exclusive access to the compute unit until it has finished execution,
after which a subsequent workgroup can be bound to the compute unit.
Experiments suggest that the occupancy-bound execution model is widely
implemented by today's GPU devices and drivers~\cite{DBLP:conf/oopsla/SorensenDBGR16}. \ADComment{For the previous cite, should we cite other papers as well as ours?}  It is
evidently a simple model for device vendors to implement efficiently.
